---
title: "pm566_HW1"
author: "Yuhong Hu"
date: "`r Sys.Date()`"
output: github_document
always_allow_html: true

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(leaflet)
library(data.table)
library(lubridate)
library(skimr)
```

# Introduction
The primary question you will answer is whether daily concentrations of PM~2.5~ (particulate matter air pollution with aerodynamic diameter less than 2.5  μm) have decreased in California over the last 15 years (from 2004 to 2019).

# Q1
Read in the data using data.table(). For each of the two datasets, check the dimensions, headers, footers, variable names and variable types. Check for any data issues, particularly in the key variable we are analyzing. Make sure you write up a summary of all of your findings.

## Read in dataset
```{r}
pm04 <- fread("~/Desktop/PhD course/pm 566/pm566_hw/HW1/ad_viz_plotval_data_2004.csv")
pm19 <- fread("~/Desktop/PhD course/pm 566/pm566_hw/HW1/ad_viz_plotval_data_2019.csv")
class(pm04)
class(pm19)
```

Both datasets were in the format of data.table or data.frame.


## Check for 2004 data

In PM~2.5~ 2004 dataset, there are 20 variables and 19233 observations. There were 8 character variables and 12 numeric variables.All the variable we are going to use is in right format and type, except for Date, which we are going to change it from character class to correct date format. No missing value were found for key variables (Date, Sitename, county,Daily Mean PM~2.5~ Concentration) we are going to use. Based on the summary statistics (using skim function), no obvious outlier was found for most of the key variables. However, according to subject matter knowledge, the concentration of PM~2.5~ cannot be under 0. Thus, we exclude observation with PM~2.5~ < 0 (N=1).The very large PM~2.5~ value (e.g., maximum 251 μg/m^3^) is still plausible, though very abnormal.

```{r}
dim(pm04)
head(pm04)
tail(pm04)
colnames(pm04)
str(pm04)
skim(pm04)

# exclude pm2.5<0
pm04 <- pm04[pm04$`Daily Mean PM2.5 Concentration`>=0,]
dim(pm04)
```

## Check for 2019 data

In PM~2.5~ 2019 dataset, there are 20 variables and 53156 observations. There were 8 character variables and 12 numeric variables.All the variable we are going to use is in right format and type, except for Date, which we are going to change it from character class to correct date format. No missing value were found for key variables (Date, Sitename, county,Daily Mean PM~2.5~ concentration) we are going to use.Based on the summary statistics, no obvious outlier was found.Based on the summary statistics (using skim function), no obvious outlier was found for most of the key variables. However, according to subject matter knowledge, the concentration of PM~2.5~ cannot be under 0. Thus, we exclude observation with PM~2.5~ < 0 (N=282).

```{r}
dim(pm19)
head(pm19)
tail(pm19)
colnames(pm19)
str(pm19)
skim(pm19)

# exclude pm2.5<0
pm19 <- pm19[pm19$`Daily Mean PM2.5 Concentration`>=0,]
dim(pm19)
```


# Q2

Combine the two years of data into one data frame. Use the Date variable to create a new column for year, which will serve as an identifier. Change the names of the key variables so that they are easier to refer to in your code.

```{r}
pm04$year <- year(as.Date(pm04$Date,"%m/%d/%Y"))
pm19$year <- year(as.Date(pm19$Date,"%m/%d/%Y"))
# check whether the conversion was right
table(pm04$year)
table(pm19$year)

# Combine the two datasets
pm <- rbind(pm04,pm19)
dim(pm) # The number of observations = sum of the two datasets

# Rename some variables for convenience
pm <- pm %>%
  rename(pm25 = `Daily Mean PM2.5 Concentration`,
         countycode = COUNTY_CODE,
         county = COUNTY,
         site = `Site Name`,
         lat = SITE_LATITUDE,
         lon = SITE_LONGITUDE,
         )

```

# Q3

Create a basic map in leaflet() that shows the locations of the sites (make sure to use different colors for each year). Summarize the spatial distribution of the monitoring sites.
```{r}
#update the two datasets 
pm04 <- pm %>%
  filter(year==2004)
pm19 <- pm %>%
  filter(year==2019)
# create unique sites for the two years
map04 <- (unique(pm04[,c("lat","lon")])) 
map19 <- (unique(pm19[,c("lat","lon")])) 

leaflet() %>% 
  addProviderTiles('CartoDB.Positron') %>% 
  addCircles(
    data = map04,
    lat = ~lat, lng = ~lon, popup = "2004",
    opacity = 1, fillOpacity = 1, radius = 300,color = 'blue'
    ) %>%
  addCircles(
    data = map19,
    lat = ~lat, lng = ~lon, popup = "2019",
    opacity=0.7, fillOpacity=0.7, radius = 300,color = 'orange'
    )


```

As the monitor sites map shown above (sites of 2004 in blue, sites of 2019 in orange). The sites were concentrated in two big cites (Los Angles, and San Francisco) and their surroundings. The low density of sites were shown in some area close to state border (California-Arizona, California-Nevada, California-Oregon).Sites in 2019 covered almost all the monitor sites in 2004, and there were more sites put in use in 2019 compared to 2004. A very small proportion of sites in 2004 shut in 2019.

# Q4
Check for any missing or implausible values of PM~2.5~ in the combined dataset. Explore the proportions of each and provide a summary of any temporal patterns you see in these observations.

We have explored NA and implausible value in Q1, here we would focus on how the number of sites and counties with monitoring PM~2.5~ data change over time.
```{r}

pm %>%
  filter(year==2004) %>%
  distinct(county) %>%
  nrow

pm %>%
  filter(year==2019) %>%
  distinct(county) %>%
  nrow

county19 <- unique(pm[pm$year==2019,'county'])
county04 <- unique(pm[pm$year==2004,'county'])
county19$county[!(county19$county %in% county04$county)]
```
Compared to 2019 with 51 counties monitored, monitoring data from these four counties (Glenn, Madera, Napa, Tehama) were missing for 2004.

```{r}
pm %>%
  filter(year==2004) %>%
  distinct(site) %>%
  nrow

pm %>%
  filter(year==2019) %>%
  distinct(site) %>%
  nrow

site19 <- unique(pm[pm$year==2019,'site'])
site04 <- unique(pm[pm$year==2004,'site'])
```
Compared to 2019 with 159 monitoring sites, there were only 105 monitoring sites in 2004. 
27 sites in 2004 were abandoned in 2019. See below.
```{r}
site04$site[!(site04$site %in% site19$site)]
```
81 new sites were put into use in 2019 compared to 2004. See below.
```{r}
site19$site[!(site19$site %in% site04$site)]
```

Therefore, in general, monitoring data in 2019 covered more area than 2004. We also explored the temporal patterns of daily PM2.5 combining obervations from all sites.

```{r}
pm$Date <- as.Date(pm$Date,"%m/%d/%Y")
pm$yday <- yday(pm$Date)

ggplot(pm, aes(x=yday, y=pm25,color=as.factor(year))) +
  geom_line() +
  labs(title="Temporal patterns of Daily PM2.5 in all sites",
        x ="Day of the year", y = "PM2.5 (ug/m3)",
        color = "Year")
```

For all the observations, we observed in both 2004 and 2019, the PM~2.5~ level fluctuates all year around. The relative low levels were found in spring. The relative high levels were found in winter and some periods of summer and fall.

# Q5
Explore the main question of interest at three different spatial levels. Create exploratory plots (e.g. boxplots, histograms, line plots) and summary statistics that best suit each level of data. Be sure to write up explanations of what you observe in these data.

### State Level
```{r warning=FALSE}
statelevel <- pm %>%
  group_by(yday,year) %>%
  summarise(pm25av=mean(pm25))

ggplot(statelevel, aes(x=yday, y=pm25av,color=as.factor(year))) +
  geom_line() +
  labs(title="Temporal patterns of Average Daily PM2.5",
        x ="Day of the year", y = "PM2.5 (ug/m3)",
        color = "Year")
```

First, we averaged the daily PM~2.5~ across all monitoring sites in CA for each day in 2004 and 2019, and plotted the average daily level on day of the year. By comparing the temporal patterns of daily PM~2.5~ level in 2004 and 2019, we observed a decrease in PM~2.5~ from 2004 to 2019 almost in each season at state level.

```{r warning=FALSE}
pm %>%
  group_by(year) %>%
  summarise(mean=mean(pm25),
            percentile25 = quantile(pm25,probs=0.25),
            median = median(pm25),
            percentile75 = quantile(pm25,probs=0.75),
            min = min(pm25),
            max = max(pm25)
            )

```

Also, by comparing mean, median, 25th percentile,75th percentile, minimum and maximum of daily PM~2.5~ combining all sites in these two years, we observed a decrease in PM~2.5~ from 2004 to 2019 at state level.

### County Level
```{r fig.height=30, fig.width=10}
length(unique(pm$countycode))
# 51 different counties, not suitable for time-series plots
ggplot(pm,aes(x=county,y=pm25,color=as.factor(year)))+
  geom_boxplot()+
  labs(title="Daily PM2.5 distribution by county",
        x ="County", y = "PM2.5 (ug/m3)",
        color = "Year")+
  coord_flip()

pm %>%
  group_by(county,year) %>%
  summarise(mean=mean(pm25),
            percentile25 = quantile(pm25,probs=0.25),
            median = median(pm25),
            percentile75 = quantile(pm25,probs=0.75),
            min = min(pm25),
            max = max(pm25)
            )
  
```
By comparing the overall distribution of daily PM~2.5~ level boxplots for each county (across whole year monitoring data) in 2004 and 2019, and by comparing mean, median, 25th percentile,75th percentile of daily PM~2.5~ in each county (across whole year monitoring data) in these two years, we observed a decrease in daily PM~2.5~ from 2004 to 2019 in almost all the counties in California with just a few exceptions.

### Site (in Los Angeles) Level
```{r fig.height=10, fig.width=10}
pmla <- pm[pm$county=='Los Angeles']
ggplot(pmla,aes(x=site,y=pm25,color=as.factor(year)))+
  geom_boxplot()+
  labs(title="Daily PM2.5 distribution by sites in LA",
        x ="Site", y = "PM2.5 (μg/m3)",
        color = "Year")+
  coord_flip()

pmla %>%
  group_by(site,year) %>%
  summarise(mean=mean(pm25),
            percentile25 = quantile(pm25,probs=0.25),
            median = median(pm25),
            percentile75 = quantile(pm25,probs=0.75),
            min = min(pm25),
            max = max(pm25)
            )
```
By comparing the overall distribution of daily PM~2.5~ level boxplots (across whole year monitoring data) for each monitoring sites in 2004 and 2019, and by comparing mean, median, 25th percentile,75th percentile of daily PM~2.5~ in each site in LA in these two years, we observed a decrease in daily PM~2.5~ from 2004 to 2019 in each site in LA.
